---
title: "Datathon 2024 (Health): Drug Overdose in USA "
output: 
  flexdashboard::flex_dashboard:
    theme:
      version: 4
      bootswatch: lux
    orientation: columns
    social: menu
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(cache = TRUE)
library(flexdashboard)
library(readr)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(shiny)
library(highcharter)
```

```{r}
knitr::opts_chunk$set(cache = TRUE)

thm <- 
  hc_theme(
    colors = c("#1a6ecc", "black", "#90ed7d"),
    chart = list(
      backgroundColor = "transparent",
      style = list(fontFamily = "Helvetica")
    ),
    xAxis = list(
      gridLineWidth = 1
    )
  )

data_og = read.csv("Drug_overdose_death_rates__by_drug_type__sex__age__race__and_Hispanic_origin__United_States_20240518.csv")

data_crude = data_og %>%
# Using the "Deaths per 100,000 resident population, crude" rates instead of the age-adjusted rates to look at the age distributions.
  filter(UNIT == "Deaths per 100,000 resident population, crude")

### Data for age
health_data_og = read.csv("Drug_overdose_death_rates__by_drug_type__sex__age__race__and_Hispanic_origin__United_States_20240518.csv")

### Cleans up the data set
health_data = data_og %>%
  na.omit() %>%
  select(-INDICATOR, -UNIT, -STUB_NAME, -FLAG) 

### indexes_all_age = grep("All ages", health_data$AGE)

indexes_all_age = which(health_data$AGE == "All ages")

n = length(health_data$AGE)

for(i in 1:n) {
  if(i %in% indexes_all_age) {health_data$AGE[i] = NA}
}

health_data_age = health_data %>%
  na.omit()


### Filtered data 2 
### Cleans up the data set
health_data = data_og %>%
  na.omit() %>%
  select(-INDICATOR, -UNIT, -STUB_NAME, -FLAG)

### Sex
index_vector = c(1:n)

indexes_male = grep("Male", health_data$STUB_LABEL)
indexes_female = grep("Female", health_data$STUB_LABEL)
indexes_wo_sex = setdiff(index_vector, union(indexes_female, indexes_male))


for(i in 1:n) {
  health_data$SEX[i] = ifelse(i %in% indexes_wo_sex, NA,
                           ifelse(i %in% indexes_male, "Male",
                                  ifelse(i %in% indexes_female, "Female", NA)))
}

### Race
indexes_white = grep("White", health_data$STUB_LABEL)
indexes_black = grep("Black", health_data$STUB_LABEL)
indexes_asian = grep("Asian", health_data$STUB_LABEL)
indexes_hs_lat = grep("All races", health_data$STUB_LABEL)
indexes_native = grep("American Indian or Alaska Native", health_data$STUB_LABEL)
indexes_island = grep("Islander", health_data$STUB_LABEL)

for(i in 1:n) {
  health_data$RACE[i] = ifelse(i %in% indexes_white, "White",
                           ifelse(i %in% indexes_black, "Black",
                                  ifelse(i %in% indexes_asian, "Asian", 
                                         ifelse(i %in% indexes_hs_lat, "Hispanic or Latino",
                                                ifelse(i %in% indexes_native, "Native American",
                                                       ifelse(i %in% indexes_island, "Islander", NA))))))
}

```





Data By Demographic
================================

Questions to Explore: {.sidebar}
-------------------------------------

1. What are the trends in drug overdose death rates over time?

```{r}
sliderInput("year0", label = "Year:",
            min = 1999, max = 2018, value = c(1999,2018), step = 1, ticks=F, sep="")
selectInput("layer", label = "Deaths By:",
            choices = c("Total Overdoses", "Any Opioid", "Natural and Semisynthetic Opioids", "Methadone", "Synthetic Opioids (other than methadone)", "Heroin"), selected = "Total Overdoses")
```

Row {.tabset}
-------------------------------------

### Drug Overdose By Age and Sex $^a$

```{r, cache=FALSE}
renderPlot({ 
  if (input$layer == "Total Overdoses") {
    panel_selection = data_crude %>%
    filter(PANEL == "All drug overdose deaths")
  } else if (input$layer == "Any Opioid") {
    panel_selection = data_crude %>%
    filter(PANEL == "Drug overdose deaths involving any opioid")
  } else if (input$layer == "Natural and Semisynthetic Opioids") {
    panel_selection = data_crude %>%
    filter(PANEL == "Drug overdose deaths involving natural and semisynthetic opioids")
  } else if (input$layer == "Methadone") {
    panel_selection = data_crude %>%
    filter(PANEL == "Drug overdose deaths involving methadone")
  } else if (input$layer == "Synthetic Opioids (other than methadone)") {
    panel_selection = data_crude %>%
    filter(PANEL == "Drug overdose deaths involving other synthetic opioids (other than methadone)")
  } else if (input$layer == "Heroin") {
    panel_selection = data_crude %>%
    filter(PANEL == "Drug overdose deaths involving heroin")}
  
  data_1 = panel_selection %>%
  # Replace NA's with 0
  filter(YEAR <= input$year0[2]) %>%
  filter(YEAR >= input$year0[1]) %>%
  mutate(EST = ifelse(is.na(ESTIMATE), 0, ESTIMATE)) %>%
  group_by(STUB_LABEL_NUM, STUB_LABEL, PANEL_NUM) %>%
  summarise(EST = sum(EST))
  
ggplot(data_1, aes(x=reorder(STUB_LABEL, EST), y = EST)) + 
  geom_bar(stat = "identity")+
  coord_flip() +
  ylab("Estimate (Deaths Per 100,000 summed over the years selected)")+
  xlab("Age and Sex Group")
})

```

### Drug Overdose Over Time By Age $^a$ 

```{r, cache=FALSE}
renderPlot({ 
  if (input$layer == "Total Overdoses") {
    panel_selection = data_crude %>%
    filter(PANEL == "All drug overdose deaths")
  } else if (input$layer == "Any Opioid") {
    panel_selection = data_crude %>%
    filter(PANEL == "Drug overdose deaths involving any opioid")
  } else if (input$layer == "Natural and Semisynthetic Opioids") {
    panel_selection = data_crude %>%
    filter(PANEL == "Drug overdose deaths involving natural and semisynthetic opioids")
  } else if (input$layer == "Methadone") {
    panel_selection = data_crude %>%
    filter(PANEL == "Drug overdose deaths involving methadone")
  } else if (input$layer == "Synthetic Opioids (other than methadone)") {
    panel_selection = data_crude %>%
    filter(PANEL == "Drug overdose deaths involving other synthetic opioids (other than methadone)")
  } else if (input$layer == "Heroin") {
    panel_selection = data_crude %>%
    filter(PANEL == "Drug overdose deaths involving heroin")}
data_2 = panel_selection %>%
  # Filter year ranges here:
  filter(YEAR <= input$year0[2]) %>%
  filter(YEAR >= input$year0[1]) %>%
  filter(STUB_NAME == "Age") %>%
  # Replace NA's with 0
  mutate(EST = ifelse(is.na(ESTIMATE), 0, ESTIMATE)) %>%
  group_by(AGE, YEAR) %>%
  summarise(EST = sum(EST))

ggplot(data_2, aes(x=YEAR, y = AGE, fill = EST))+
  geom_tile(color = "white")+
  scale_fill_gradient(low = "lightblue", high = "darkred") +
  ylab("Age Group")+
  xlab("Year")+
  labs(fill = "Estimate \n (Deaths Per 100,000)")
})
```

### By Race $^b$

```{r, cache=FALSE}
renderPlot({ 
  if (input$layer == "Total Overdoses") {
    panel_selection = health_data %>%
    filter(PANEL == "All drug overdose deaths")
  } else if (input$layer == "Any Opioid") {
    panel_selection = health_data %>%
    filter(PANEL == "Drug overdose deaths involving any opioid")
  } else if (input$layer == "Natural and Semisynthetic Opioids") {
    panel_selection = health_data %>%
    filter(PANEL == "Drug overdose deaths involving natural and semisynthetic opioids")
  } else if (input$layer == "Methadone") {
    panel_selection = health_data %>%
    filter(PANEL == "Drug overdose deaths involving methadone")
  } else if (input$layer == "Synthetic Opioids (other than methadone)") {
    panel_selection = health_data %>%
    filter(PANEL == "Drug overdose deaths involving other synthetic opioids (other than methadone)")
  } else if (input$layer == "Heroin") {
    panel_selection = health_data %>%
    filter(PANEL == "Drug overdose deaths involving heroin")}
  
  data_3 = panel_selection %>%
    filter(YEAR <= input$year0[2]) %>%
    filter(YEAR >= input$year0[1]) %>%
    na.omit %>%
    group_by(RACE, PANEL_NUM) %>%
    summarise(ESTIMATE = sum(ESTIMATE))
  
ggplot(data_3, aes(x=reorder(RACE, ESTIMATE), y = ESTIMATE)) + 
  geom_bar(stat = "identity") +
  coord_flip()+
  ylab("Estimate (Deaths Per 100,000)")+
  xlab("Race")
})
```

### By Age $^b$

```{r, cache=FALSE}
renderPlot({ 
  if (input$layer == "Total Overdoses") {
    panel_selection = health_data_age %>%
    filter(PANEL == "All drug overdose deaths")
  } else if (input$layer == "Any Opioid") {
    panel_selection = health_data_age %>%
    filter(PANEL == "Drug overdose deaths involving any opioid")
  } else if (input$layer == "Natural and Semisynthetic Opioids") {
    panel_selection = health_data_age %>%
    filter(PANEL == "Drug overdose deaths involving natural and semisynthetic opioids")
  } else if (input$layer == "Methadone") {
    panel_selection = health_data_age %>%
    filter(PANEL == "Drug overdose deaths involving methadone")
  } else if (input$layer == "Synthetic Opioids (other than methadone)") {
    panel_selection = health_data_age %>%
    filter(PANEL == "Drug overdose deaths involving other synthetic opioids (other than methadone)")
  } else if (input$layer == "Heroin") {
    panel_selection = health_data_age %>%
    filter(PANEL == "Drug overdose deaths involving heroin")}
  
  data_4 = panel_selection %>%
    filter(YEAR <= input$year0[2]) %>%
    filter(YEAR >= input$year0[1]) %>%
    group_by(AGE, PANEL_NUM) %>%
    summarise(ESTIMATE = sum(ESTIMATE))
  
ggplot(data_4, aes(x=reorder(AGE, ESTIMATE), y = ESTIMATE)) + 
  geom_bar(stat = "identity") +
  coord_flip()+
  ylab("Estimate (Deaths Per 100,000)")+
  xlab("Age Group")
})

```

### By Gender $^b$

```{r, cache=FALSE}
renderPlot({ 
  if (input$layer == "Total Overdoses") {
    panel_selection = health_data %>%
    filter(PANEL == "All drug overdose deaths")
  } else if (input$layer == "Any Opioid") {
    panel_selection = health_data %>%
    filter(PANEL == "Drug overdose deaths involving any opioid")
  } else if (input$layer == "Natural and Semisynthetic Opioids") {
    panel_selection = health_data %>%
    filter(PANEL == "Drug overdose deaths involving natural and semisynthetic opioids")
  } else if (input$layer == "Methadone") {
    panel_selection = health_data %>%
    filter(PANEL == "Drug overdose deaths involving methadone")
  } else if (input$layer == "Synthetic Opioids (other than methadone)") {
    panel_selection = health_data %>%
    filter(PANEL == "Drug overdose deaths involving other synthetic opioids (other than methadone)")
  } else if (input$layer == "Heroin") {
    panel_selection = health_data %>%
    filter(PANEL == "Drug overdose deaths involving heroin")}
  
  data_5 = panel_selection %>%
    filter(YEAR <= input$year0[2]) %>%
    filter(YEAR >= input$year0[1]) %>%
    na.omit %>%
    group_by(SEX, PANEL_NUM) %>%
    summarise(ESTIMATE = sum(ESTIMATE))
  
  ggplot(data_5, aes(x=SEX, y = ESTIMATE)) + 
    geom_bar(stat = "identity", width = 0.4) +
    ylab("Estimate (Deaths Per 100,000)") +
    xlab("Sex")
})
```

### Methodology For Data Visualizations: $^{a \ and \ b}$

Method a:

* NA's were replaced with 0
* Using the "Deaths per 100,000 resident population, crude"  instead of the age-adjusted rates
* Otherwise the same as the data set given


Method b:

* Reordered the data set
* NA's were ommitted

Analysis
================================


Machine Learning Model
================================
