---
title: "Machine Learning Model"
author: "Stat342 Team"
date: "2024-05-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyverse)
library(VIM)
library(dplyr)
library(skimr)
library(mice)
library(caret)
library(randomForest)
library(xgboost)
library(e1071)
```


```{r}
data <- read.csv("Drug_overdose_death_rates__by_drug_type__sex__age__race__and_Hispanic_origin__United_States_20240518.csv")
skim(data)

# Visualize the pattern of missing values
aggr(data, col = c('navyblue', 'red'), numbers = TRUE, sortVars = TRUE, labels = names(data), cex.axis = 0.7, gap = 3, ylab = c("Missing data", "Pattern"))
```
We see there is no duplicate rows and 'Estimate' variable has 1111 missing values, which is 17.84% of the total data.

# Data Cleaning:

We already have some numeric variables for levels of category variables, so we decided to delete the category variables. Also, we dropped the Flag variable since it's not critical for analysis. For the missing value in Estimate, we replace them by using Multiple Imputation.

```{r,include=FALSE}
data <- data %>% select(-INDICATOR, -PANEL, -UNIT, -STUB_NAME, -STUB_LABEL, -YEAR, -AGE, -FLAG)

#Multiple Imputation using mice
imputed_data <- mice(data, m = 5, method = 'pmm', seed = 500)
complete_data <- complete(imputed_data, 1)
```

# Partition data into training and test datasets:

```{r}
set.seed(123)

# Partition data into training and test sets
trainIndex <- createDataPartition(complete_data$ESTIMATE, p = 0.8, list = FALSE, times = 1)
dataTrain <- complete_data[trainIndex, ]
dataTest <- complete_data[-trainIndex, ]

# Check the dimensions of the resulting datasets
dim(dataTrain)
dim(dataTest)
```

# Train Random Forest Model:

```{r}
RF_model <- randomForest(ESTIMATE ~ ., data = dataTrain)
RF_model
```
An explained variance of 86.26% suggests that the model is performing well and is able to explain a large proportion of the variability in drug overdose death rates. 

# Evaluate Random Forest Model:

```{r}
predicted <- predict(RF_model, dataTest)

# Calculate RMSE
rmse <- sqrt(mean((dataTest$ESTIMATE - predicted)^2))

# Calculate 95% confidence interval for RMSE
n <- nrow(dataTest) # Number of observations in test data
t_value <- qt(0.975, df = n - 1) # t-value for 95% confidence interval
margin_of_error <- t_value * (rmse / sqrt(n))

# Confidence interval
lower_bound <- rmse - margin_of_error
upper_bound <- rmse + margin_of_error

# Print results
print(paste("RMSE:", rmse))
print(paste("Marginal of Error:",margin_of_error))
print(paste("95% Confidence Interval for RMSE:", lower_bound, "-", upper_bound))
```

In this case, an RMSE of approximately 2.057 indicates that, on average, the model's predictions are off by about 2.057 units of the dependent variable (drug overdose death rates). Besides, with a 95% confidence level, we expect that if we were to repeat this process multiple times, approximately 95% of the time, the true RMSE would fall within the range of 1.943 to 2.172.

# Hyperparameter Tuning

To further assess the model’s performance, we tried experiments with different hyperparameters, such as the number of trees and the number of variables tried at each split, to see if the model performance can be improved.

```{r}
# Define the grid of hyperparameters to search
tune_grid <- expand.grid(
  mtry = c(2, 3, 4, 5)  # Number of variables randomly sampled as candidates at each split
)

# Set up cross-validation
train_control <- trainControl(method = "cv", number = 5, search = "grid")

# Train the Random Forest model with hyperparameter tuning
set.seed(123)
tuned_rf_model <- train(
  ESTIMATE ~ PANEL_NUM + UNIT_NUM + STUB_NAME_NUM + STUB_LABEL_NUM + YEAR_NUM + AGE_NUM, 
  data = dataTrain,
  method = "rf",
  trControl = train_control,
  tuneGrid = tune_grid,
  ntree = 500
)

# Print the best hyperparameters
print(tuned_rf_model$bestTune)
```
The output indicates that the best value for the mtry parameter, as determined by the hyperparameter tuning process, is 5.

# Evaluate the Tuned Model
```{r}
# Predict on the test data using the best model
predictions <- predict(tuned_rf_model, dataTest)

# Calculate RMSE
rmse <- sqrt(mean((dataTest$ESTIMATE - predictions)^2))

# Calculate 95% confidence interval for RMSE
n <- nrow(dataTest) 
t_value <- qt(0.975, df = n - 1) 
margin_of_error <- t_value * (rmse / sqrt(n))

# Confidence interval
lower_bound <- rmse - margin_of_error
upper_bound <- rmse + margin_of_error

# Print results
cat("RMSE:", rmse, "\n")
print(paste("Marginal of Error:",margin_of_error))
cat("95% Confidence Interval for RMSE:", lower_bound, "-", upper_bound, "\n")
```

# Exploring feature importance in a Random Forest:

```{r}
importance <- varImp(tuned_rf_model)
plot(importance, main = "Variable Importance for Predicting Drug Overdose Death Rates")
```
Panel demographic factors have the most influence on drug overdose death rates.




# Comparing with XGBoost Model:

```{r,warning=FALSE}
# Convert data to DMatrix format for XGBoost
train_matrix <- xgb.DMatrix(as.matrix(dataTrain[, -7]), label = dataTrain$ESTIMATE)
test_matrix <- xgb.DMatrix(as.matrix(dataTrain[, -7]))

# Train XGBoost model
xgb_model <- xgboost(data = train_matrix, objective = "reg:squarederror", nrounds = 100)
xgb_model
```

```{r,warning=FALSE}
# Evaluate XGBoost model
predicted_xgb <- predict(xgb_model, test_matrix)

# Calculate RMSE
rmse_xgb <- sqrt(mean((dataTest$ESTIMATE - predicted_xgb)^2))

# Calculate 95% confidence interval for RMSE
n <- nrow(dataTest)
t_value <- qt(0.975, df = n - 1)
margin_of_error_xgb <- t_value * (rmse_xgb / sqrt(n))

# Confidence interval
lower_bound_xgb <- rmse_xgb - margin_of_error_xgb
upper_bound_xgb <- rmse_xgb + margin_of_error_xgb

# Print results
print(paste("XGBoost RMSE:", rmse_xgb))
print(paste("Marginal of Error:",margin_of_error_xgb))
print(paste("XGBoost 95% Confidence Interval for RMSE:", lower_bound_xgb, "-", upper_bound_xgb))
```
We see the Random Forest model appears to be more stable and consistent in its predictions, as evidenced by its lower RMSE and narrower confidence interval.
